Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-10 MScDiplom_Lemenkova.tex
added hyperreferences from pictures to text; changes in 'In-situ observations', p.15, added pictuers of SCUBA and sonar

	\acro{6SV1}{Second Simulation of a Satellite Signal in the Solar Spectrum Version1} 

@@

	\acro{GPS}{Global Positioning System}

@@

@@

	\acro{UML}{Unified Modelling Language}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses\index[concepts]{Seagrasses} create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

\begin{figure}[h]
	\centering
	\subfloat {\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-1-1a.jpg}}
	\hspace{2mm}
	\subfloat {\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-1-1b.jpg}}
	\caption{Seagrass \textit{Posidonia Oceanica}}
	\label{fig:1}
\end{figure}

and is widely spread along the coasts of Crete \cite{Dumay02} \label{Dumay02}. 
It plays an important role in a number
of geomorphological and ecological processes. Namely, it is a source of food for herbivorous fauna as
well as shelter zones for fish and other marine organisms; it contributes to the nutrient recycling; it
provides sediments stability by reducing the degree of water movements, etc \cite{Francour99}\label{Francour99}.\\
The purpose of current MSc research work aims to apply methods of remote sensing analysis, including \ac{RTM}, 
\ac{GIS}-based spatial analysis, processing and classification of satellite and aerial photos, as well as videometric underwater footage, towards mapping and environmental monitoring of seagrass \textit{P. oceanica} along the selected locations of the northern coasts of Crete island, Greece. The technical implementation is based on \ac{WASI} \ac{RTM} software, \ac{GIS} (\ac{ILWIS}, Erdas Imagine and ArcGIS), using aerial and satellite images and the results of the underwater videometric measurements.

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species\index[concepts]{Seagrasses!global species} (Fig. \ref{fig:2}), belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

\begin{wrapfigure}{r}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.40]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world. Source:\cite{Green03}\label{Green03} 
	\label{fig:2}} 
\end{wrapfigure}

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas\index[concepts]{Seagrasses!distribution of}), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species \index[concepts]{Seagrasses!Mediterranean species} in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

\subsubsection{Ecological significance of the seagrasses}
Seagrass plays vital role in the marine ecosystems of the world ocean. 
Seagrasses are the only flowering plant in the world that is able to live completely submerged. Seagrass is a habitat for
numerous marine fish species \cite{Nagelkerken00}\label{Nagelkerken00}, source of primary production and food for
fish, turtles and other organisms, which gives them special environmental value \cite{Noralez10}\label{Noralez10}.\\
Seagrass meadows produce enormous quantities of organic matter (leaves, epiphytes), which
constitutes the basis of the food web both within and outside the ecosystem \cite{Gobert06}\label{Gobert06}.
Finally, seagrasses are an important component in the environmental ``food chain" of the coastal

\begin{wrapfigure}{r}{0.35\textwidth}
	\centering
	\includegraphics[scale=0.3]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}

ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine ``\textit{algae}"\index[concepts]{Algae}, ``\textit{seagrasses}" are vastly different from them.
There are fundamental differences between both marine organisms, which we briefly mention below: 
\begin{itemize}
	\item[*] first, \textit{seagrasses} are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas \textit{algae} are simple organisms that can only holdfast; 
	\item[*] secondly, \textit{seagrasses} are complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while \textit{algae} are primitive organisms with simple cell structure, spores and gametes; 
	\item[*] finally, \textit{seagrasses} uptake nutrient through the root system, while \textit{algae} nourish directly from the water column \cite{Dixon05}\label{Dixon05}. 
\end{itemize}
There are other nuances and details of difference between the ``\textit{seagrasses}" and ``\textit{algae}", but their discussion would go beyond the scope of this work.

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. 
One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}. 
The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-4.jpg}
	\caption{Distribution of seagrass in relation to mean ocean temperature. \\
	Source: \cite{Orth06}\label{Orth06}}
	\label{Fig.4}
\end{wrapfigure}

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subjectto catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 \% of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.
\pagebreak

\subsubsection{General characteristics of \textit{P.oceanica}}\index[concepts]{Seagrasses!\textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.20]{Fig-1-5.jpg}
	\caption{Geographical distribution of
	\textit{P.oceanica}. Source: \cite{Borum04}\label{Borum04}}
	\label{fig:5}
\end{wrapfigure}

The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area\index[concepts]{Seagrasses!\textit{P.oceanica}!distribution} (Fig.\ref{fig:5}), 
with its extension limited by the western part of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphologically\index[concepts]{Seagrasses!\textit{P.oceanica}!morphology}, \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots and short rhizomes (0.5-2.0 mm). 
The leaves are, perhaps, the most particular characteristics of \textit{P.oceanica}, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:8}(a)): having usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}(c)).
Growing \textit{P.oceanica} make a meadows which, in turn, consist of smaller patches, called \textit{matte}
(Fig.\ref{fig:8}(b)), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.

\begin{figure}[H]
	\centering
	\subfloat [\textit{P.oceanica}. Source \cite{Luque04}\label{Luque04}]{\includegraphics[width=0.25\textwidth]{Fig-1-6.jpg}}
	\subfloat [Scheme of matte structure of \textit{P.oceanica}. Source\cite{Pergent90}\label{Pergent90}]{\includegraphics[width=0.3\textwidth]{Fig-1-7.jpg}}
	\subfloat [Structure and components of \textit{P.oceanica}. Source\cite{DiCarlo04}\label{DiCarlo04}]{\includegraphics[width=0.3\textwidth]{Fig-1-8.jpg}}
	\caption{Morphology of \textit{P.oceanica}}\label{fig:8}
\end{figure}

@@

\begin{wrapfigure}{r}{0.3\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-9.jpg}
	\caption{Spectra of the seagrass on different depths (0 - 15 m). Source \cite{Edwards00}\label{Edwards00}}
	\label{fig:9}
\end{wrapfigure}

@@

research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring. 
The general overview of the application of various remote sensing data types (colour,

@@

have been reported by several authors \cite{Kelly80,Walker89,Green96}\label{Kelly80} \label{Walker89} \label{Green96}.

@@

seagrass mapping \cite{Calvo03,Dekker05a,Fornes06,Green96,Jackson07,Jensen95,Kendrick00,Lyzenga81,Malthus03,Matarrese08,Mount03,Pasqualini01,Pasqualini99,Pasqualini98a,Pergent-Martini06,Ralph05,Ribed02,Short01,Walker89}\label{Calvo03}\label{Dekker05a}\label{Fornes06}\label{Green96}\label{Jackson07}\label{Jensen95}\label{Kendrick00} \label{Lyzenga81}\label{Malthus03}\label{Matarrese08}\label{Mount03}\label{Pasqualini01}\label{Pasqualini99}\label{Pasqualini98a}\label{Pergent-Martini06}\label{Ralph05}\label{Ribed02}\label{Short01}\label{Walker89}. The application of the remote sensing data towards seagrass mapping is based on the spectral reflectance
 characteristics of the \textit{P.oceanica} seagrass, which enable its spectral discrimination from spectra of other seafloor types. 

@@

values \cite{Fornes06} \label{Fornes06}, so that the seafloor can be divided into several types: sand, rock, P.
oceanica, other vegetation, etc.

@@

recommendations are being elaborated. The development of the underwater SCUBA diving

>>>>

	\acro{6SV1}{\href{http://6s.ltdri.org/}{Second Simulation of a Satellite Signal in the Solar Spectrum Version1}} 

@@

	\acro{GPS}{\href{http://en.wikipedia.org/wiki/GPS}{Global Positioning System}}

@@

	\acro{LAD}{\href{http://en.wikipedia.org/wiki/Least_absolute_deviations}{Least Absolute Deviations}}

@@

	\acro{UML}{\href{http://en.wikipedia.org/wiki/Unified_Modeling_Language}{Unified Modelling Language}}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses\index[concepts]{Seagrasses} create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses (Fig.\ref{fig:2}),
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig.\ref{fig:1}\footnote{The page where the illustration is explained in the main text of MSc thesis is given in angle brackets (e.g., Page: $<$\pageref{page-40}$>$}\label{page-1}

\begin{figure}[H]
	\centering
	\subfloat {\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-1-1a.jpg}}
	\hspace{2mm}
	\subfloat {\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-1-1b.jpg}}
	\caption{Seagrass \textit{Posidonia Oceanica}. Pages: $<$\pageref{page-1}$>$, $<$\pageref{page-3}$>$, $<$\pageref{page-5}$>$}
	\label{fig:1}
\end{figure}

and is widely spread along the coasts of Crete \cite{Dumay02} \label{Dumay02}. 
It plays an important role in a number
of geomorphological and ecological processes. Namely, it is a source of food for herbivorous fauna as
well as shelter zones for fish and other marine organisms; it contributes to the nutrient recycling; it
provides sediments stability by reducing the degree of water movements, etc \cite{Francour99}\label{Francour99}.\\
The purpose of current MSc research work aims to apply methods of remote sensing analysis, including \ac{RTM}, 
\ac{GIS}-based spatial analysis, processing and classification of satellite and aerial photos, as well as videometric underwater footage, towards mapping and environmental monitoring of seagrass \textit{P. oceanica} along the selected locations of the northern coasts of Crete island, Greece (Fig.\ref{fig:3.1})\label{page-2}. The technical implementation is based on \ac{WASI} \ac{RTM} software, \ac{GIS} (\ac{ILWIS}, Erdas Imagine and ArcGIS), using aerial and satellite images and the results of the underwater videometric measurements.

\subsection{Background}\label{page-2}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species\index[concepts]{Seagrasses!global species} (Fig.\ref{fig:2}, Fig.\ref{Fig.4})\label{page-2}, belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

\begin{wrapfigure}{r}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.40]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world. Source:\cite{Green03}\label{Green03}. Pages: $<$\pageref{page-1}$>$, $<$\pageref{page-2}$>$}\label{fig:2}
\end{wrapfigure}

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas, Fig.\ref{Fig.4}\index[concepts]{Seagrasses!distribution of}), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species \index[concepts]{Seagrasses!Mediterranean species} in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

\subsubsection{Ecological significance of the seagrasses}
Seagrass plays vital role in the marine ecosystems of the world ocean. 
Seagrasses are the only flowering plant in the world that is able to live completely submerged. Seagrass is a habitat for
numerous marine fish species \cite{Nagelkerken00}\label{Nagelkerken00}, source of primary production and food for
fish, turtles and other organisms, which gives them special environmental value \cite{Noralez10}\label{Noralez10}.\\
Seagrass meadows produce enormous quantities of organic matter (leaves, epiphytes), which
constitutes the basis of the food web both within and outside the ecosystem \cite{Gobert06}\label{Gobert06}.
Finally, seagrasses are an important component in the environmental ``food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine ``\textit{algae}"\index[concepts]{Algae}, ``\textit{seagrasses}" are vastly different from them.
There are fundamental differences between both marine organisms, which we briefly mention below: 
\begin{itemize}
	\item[*] first, \textit{seagrasses} are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas \textit{algae} are simple organisms that can only holdfast; 
	\item[*] secondly, \textit{seagrasses} are complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while \textit{algae} are primitive organisms with simple cell structure, spores and gametes; 
	\item[*] finally, \textit{seagrasses} uptake nutrient through the root system, while \textit{algae} nourish directly from the water column \cite{Dixon05}\label{Dixon05}. 
\end{itemize}
There are other nuances and details of difference between the ``\textit{seagrasses}" and ``\textit{algae}", but their discussion would go beyond the scope of this work.

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} (Fig.\ref{fig:1})\label{page-3} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. 
One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06}\label{Leoni07}. 
The detailed research of the fish farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.

\begin{wrapfigure}{r}{0.35\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}. Page: $<$\pageref{page-5}$>$}
	\label{fig:3}
\end{wrapfigure}

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature\label{page-4}; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-4.jpg}
	\caption{Distribution of seagrass in relation to mean ocean temperature. 	Source: \cite{Orth06}\label{Orth06}. Pages: $<$\pageref{page-2}$>$, $<$\pageref{page-4}$>$}
	\label{Fig.4}
\end{wrapfigure}

Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of \textit{P.oceanica} in the Mediterranean Sea (Fig.\ref{fig:5})\label{page-4} are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 \% of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.
\pagebreak

\subsubsection{General characteristics of \textit{P.oceanica}}\index[concepts]{Seagrasses!\textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows (Fig.\ref{fig:1}) in the Mediterranean Sea between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.20]{Fig-1-5.jpg}
	\caption{Geographical distribution of \textit{P.oceanica}. Source: \cite{Borum04}\label{Borum04}. Pages: $<$\pageref{page-4}$>$, $<$\pageref{page-5}$>$}
	\label{fig:5}
\end{wrapfigure}

The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area\index[concepts]{Seagrasses!\textit{P.oceanica}!distribution} (Fig.\ref{fig:5}), 
with its extension limited by the western part of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphologically\index[concepts]{Seagrasses!\textit{P.oceanica}!morphology}, \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots and short rhizomes (0.5-2.0 mm), Fig.\ref{fig:8}, Fig.\ref{fig:3}.\label{page-5}
The leaves are, perhaps, the most particular characteristics of \textit{P.oceanica}, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:8}(a)): having usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}(c)).
Growing \textit{P.oceanica} make a meadows which, in turn, consist of smaller patches, called \textit{matte}
(Fig.\ref{fig:8}(b)), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.

\begin{figure}[H]
	\centering
	\subfloat [\textit{P.oceanica}. Source \cite{Luque04}\label{Luque04}]{\includegraphics[width=0.25\textwidth]{Fig-1-6.jpg}}
	\subfloat [Scheme of matte structure of \textit{P.oceanica}. Source\cite{Pergent90}\label{Pergent90}]{\includegraphics[width=0.3\textwidth]{Fig-1-7.jpg}}
	\subfloat [Structure and components of \textit{P.oceanica}. Source\cite{DiCarlo04}\label{DiCarlo04}]{\includegraphics[width=0.3\textwidth]{Fig-1-8.jpg}}
	\caption{Morphology of \textit{P.oceanica}. Page: $<$\pageref{page-5}$>$}\label{fig:8}
\end{figure}

@@

\begin{wrapfigure}{r}{0.3\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-9.jpg}
	\caption{Spectra of the seagrass on different depths (0 - 15 m). Source \cite{Edwards00}\label{Edwards00}. Page: $<$\pageref{page-7}$>$}
	\label{fig:9}
\end{wrapfigure}

@@

research object. \\
The remote sensing techniques have traditionally been widely used for the seagrass
monitoring. The general overview of the application of various remote sensing data types (colour,

@@

have been reported by several authors \cite{Kelly80,Walker89,Green96}\label{Kelly80}\label{Walker89}\label{Green96}.

@@

seagrass mapping \cite{Calvo03,Dekker05a,Fornes06,Green96,Jackson07,Jensen95,Kendrick00,Lyzenga81,Malthus03,Matarrese08,Mount03,Pasqualini01,Pasqualini99,Pasqualini98a,Pergent-Martini06,Ralph05,Ribed02,Short01,Walker89}\label{Calvo03}\label{Dekker05a}\label{Fornes06}\label{Green96}\label{Jackson07}\label{Jensen95}\label{Kendrick00}\label{Lyzenga81}\label{Malthus03}\label{Matarrese08}\label{Mount03}\label{Pasqualini01}\label{Pasqualini99}\label{Pasqualini98a}\label{Pergent-Martini06}\label{Ralph05}\label{Ribed02}\label{Short01}\label{Walker89}. \\The application of the remote sensing data towards seagrass mapping is based on the spectral reflectance  characteristics of the \textit{P.oceanica} seagrass, which enable its spectral discrimination from spectra of other seafloor types. 

@@

values (Fig.\ref{fig:9})\cite{Fornes06}\label{Fornes06}, so that the seafloor can be divided into several types: sand, rock, \textit{P.oceanica}, other vegetation, etc\label{page-7}.

@@

recommendations are being elaborated. The development of the underwater \ac{SCUBA} diving
